---
title: "GPS Session Analysis — Physical Load Monitoring"
author: "Hector Cardenas"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Overview

GPS tracking data from two training sessions (MD-3 and MD-4) of a professional football squad. The goal is to quantify physical load by **position** and **session type**, identify the most demanding players, and flag differences between training days — the kind of insight a Sport Scientist delivers daily.

**Dataset:** 22 players · 2 sessions · 6 positions · 23 physical metrics

---

## 1. Setup & Data

```{r libraries}
library(tidyverse)
library(ggrepel)
library(scales)
library(patchwork)
```

```{r load-data}
data_gps <- read.csv("data/data_gps.csv") %>%
  rename(
    date          = Date,
    player        = Name,
    session       = Session,
    position      = Position.Name,
    duration_min  = Total.Duration,
    distance_m    = Total.Distance..m.,
    player_load   = Total.Player.Load,
    hsr_m         = High.speed.distance..m.,
    hsr_efforts   = High.speed.efforts,
    sprint_m      = Sprint.distance..m.,
    sprint_eff    = Sprints.efforts,
    max_vel_kmh   = Maximum.Velocity..km.h.
  ) %>%
  filter(distance_m > 0) %>%
  mutate(
    load_per_min   = round(player_load / duration_min, 2),
    hsr_ratio      = round(hsr_m / distance_m * 100, 1),
    session        = factor(session, levels = c("MD-4", "MD-3"))
  )

glimpse(data_gps)
```

---

## 2. Physical Demand by Position

Players at different positions cover different distances and generate different loads. We compare **total distance**, **player load**, and **high-speed running (HSR)** across all positions.

```{r position-overview, fig.height=7}
pos_summary <- data_gps %>%
  group_by(position) %>%
  summarise(
    n              = n(),
    avg_distance   = mean(distance_m),
    avg_load       = mean(player_load),
    avg_hsr        = mean(hsr_m),
    avg_max_vel    = mean(max_vel_kmh),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_distance))

p1 <- ggplot(pos_summary, aes(x = reorder(position, avg_distance), y = avg_distance, fill = position)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(avg_distance), "m")), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Average Total Distance by Position", x = NULL, y = "Distance (m)") +
  theme_minimal(base_size = 12)

p2 <- ggplot(pos_summary, aes(x = reorder(position, avg_hsr), y = avg_hsr, fill = position)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(avg_hsr), "m")), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Average HSR Distance by Position", x = NULL, y = "HSR Distance (m)") +
  theme_minimal(base_size = 12)

p1 / p2
```

**Key finding:** Wide players (EXTREMO/LATERAL) accumulate the most HSR, while central midfielders cover more total distance. This reflects position-specific demands: wide players sprint in behind, central mids circulate.

---

## 3. Session Comparison — MD-4 vs MD-3

As match day approaches, training intensity typically increases. We verify this using **player load per minute** (intensity index) and **HSR ratio** (% of total distance at high speed).

```{r session-comparison}
session_summary <- data_gps %>%
  group_by(session, position) %>%
  summarise(
    avg_load_min = mean(load_per_min),
    avg_hsr_ratio = mean(hsr_ratio),
    avg_distance = mean(distance_m),
    .groups = "drop"
  )

p3 <- ggplot(data_gps, aes(x = session, y = load_per_min, fill = session)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_manual(values = c("MD-4" = "#4e9af1", "MD-3" = "#e05c5c")) +
  facet_wrap(~position, nrow = 2) +
  labs(
    title    = "Player Load per Minute — MD-4 vs MD-3",
    subtitle = "MD-3 (closer to match) is expected to be higher intensity",
    x = NULL, y = "Player Load / min",
    fill = "Session"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p3
```

```{r hsr-session}
p4 <- ggplot(data_gps, aes(x = session, y = hsr_ratio, fill = session)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  scale_fill_manual(values = c("MD-4" = "#4e9af1", "MD-3" = "#e05c5c")) +
  labs(
    title = "% Distance at High Speed (HSR Ratio) — MD-4 vs MD-3",
    x = NULL, y = "HSR Distance / Total Distance (%)",
    fill = "Session"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

p4
```

---

## 4. Individual Player Rankings

Identifying who covered the most ground and who was most explosive across both sessions.

```{r player-ranking, fig.height=8}
player_total <- data_gps %>%
  group_by(player, position) %>%
  summarise(
    total_distance = sum(distance_m),
    total_hsr      = sum(hsr_m),
    total_sprints  = sum(sprint_eff),
    avg_max_vel    = mean(max_vel_kmh),
    .groups = "drop"
  )

top_distance <- player_total %>% slice_max(total_distance, n = 10)

p5 <- ggplot(top_distance, aes(x = reorder(player, total_distance),
                               y = total_distance, fill = position)) +
  geom_col() +
  geom_text(aes(label = paste0(round(total_distance / 1000, 1), "km")),
            hjust = -0.1, size = 3.2) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = "Top 10 Players — Total Distance (Both Sessions)",
    x = NULL, y = "Total Distance (m)", fill = "Position"
  ) +
  theme_minimal(base_size = 11)

p5
```

```{r speed-scatter}
ggplot(player_total, aes(x = total_distance, y = total_hsr, color = position)) +
  geom_point(size = 4, alpha = 0.8) +
  geom_label_repel(aes(label = player), size = 2.8, max.overlaps = 12) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title    = "Physical Profile Map — Distance vs HSR",
    subtitle = "Top-right quadrant = high volume AND high intensity",
    x = "Total Distance (m)", y = "Total HSR Distance (m)",
    color = "Position"
  ) +
  theme_minimal(base_size = 12)
```

---

## 5. Z-Score Monitoring — Who Stands Out?

Z-scores normalize each player's output against the group mean for that session and position. Values above +1.5 or below -1.5 warrant attention (overload or underperformance).

```{r zscores, fig.height=7}
z_data <- data_gps %>%
  group_by(session, position) %>%
  mutate(
    z_distance  = scale(distance_m)[,1],
    z_load      = scale(player_load)[,1],
    z_hsr       = scale(hsr_m)[,1]
  ) %>%
  ungroup()

z_long <- z_data %>%
  select(player, session, position, z_distance, z_load, z_hsr) %>%
  pivot_longer(cols = starts_with("z_"), names_to = "metric", values_to = "z") %>%
  mutate(metric = recode(metric,
    z_distance = "Total Distance",
    z_load     = "Player Load",
    z_hsr      = "HSR Distance"
  ))

ggplot(z_long %>% filter(session == "MD-3"),
       aes(x = metric, y = reorder(player, z), fill = z)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(z, 1)), size = 3) +
  scale_fill_gradient2(low = "#d73027", mid = "white", high = "#1a9850",
                       midpoint = 0, limits = c(-2.5, 2.5)) +
  labs(
    title    = "Z-Score Heatmap — MD-3 Session",
    subtitle = "Green = above group average | Red = below group average",
    x = NULL, y = NULL, fill = "Z-Score"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

---

## Summary

| Insight | Finding |
|---|---|
| Most demanding position | Wide players (EXTREMO/LATERAL) — highest HSR |
| Session intensity | MD-3 shows higher load/min vs MD-4 as expected |
| Top volume player | See ranking chart — covers most total distance |
| Monitoring flags | Z-score heatmap identifies outliers needing attention |

---

*Data: GPS tracking from professional football training sessions.*
*Analysis: R · tidyverse · ggplot2 · ggrepel · patchwork*
